A nice acyclic matching on the nerve of the partition lattice

### 2020-02-15 paper3/equivariant.tex
paper3 citation in theorems
paper3 splitting basis

\documentclass{elsarticle}
\usepackage{amsmath,amsthm}
\usepackage{bbm}
\newtheorem{df}{Definition}[section]
\newtheorem{prop}[df]{Proposition}
\newtheorem{lem}[df]{Lemma}
\newtheorem{thm}[df]{Theorem}
\newtheorem{cor}[df]{Corollary}
\newtheorem{ex}[df]{Example}
\newtheorem{rem}[df]{Remark}
\def\N{\mathbbm N}
\def\Z{\mathbbm Z}
\def\id{\operatorname{id}}
%\def\card{\operatorname{card}}
%\def\C{\mathbbm C}
\journal{arXiv.org}
\begin{document}
\begin{frontmatter}
\title{A nice acyclic matching on the nerve of the partition lattice}
\author[ruelle]{Ralf Donau}
\address[ruelle]{Fachbereich Mathematik, Universit\"at Bremen, Bibliothekstra\ss e 1, 28359 Bremen, Germany}
\ead{ruelle@math.uni-bremen.de}
\begin{abstract}
The author has already proven that the space $\Delta(\Pi_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$ for all natural numbers $n\geq 3$ and all subgroups $G\subset S_1\times S_{n-1}$. We wish to construct an acyclic matching on $\Delta(\Pi_n)/G$ that allows us to give a basis of its cohomology. This is also a more elementary approach to determining the number of spheres. Furthermore we give a description of the group action by an action on the spheres. We also obtain another result that we call Equivariant Patchwork Theorem.
\end{abstract}
\begin{keyword}
Discrete Morse Theory\sep Regular trisp\sep Acyclic matching\sep Equivariant homotopy
\end{keyword}
\end{frontmatter}
\section{Introduction}
Let $n\geq 3$ and let $\Pi_n$ denote the poset consisting of all partitions of $[n]:=\{1,\dots,n\}$ ordered by refinement, such that the finer partition is the smaller partition. Let $\overline{\Pi}_n$ denote the poset obtained from $\Pi_n$ by removing both the smallest and greatest element, which are $\{\{1\},\dots,\{n\}\}$ and $\{[n]\}$, respectively. We consider $\overline{\Pi}_n$ as a category, which is acyclic, and define $\Delta(\overline{\Pi}_n)$ to the nerve of the acyclic category $\overline{\Pi}_n$, which is a regular trisp, see \cite[Chapter 10]{buch}. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way.

It is well-known that $\Delta(\overline{\Pi}_n)$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The following two theorems are results concerning the topology of the quotient $\Delta(\overline{\Pi}_n)/G$, where $G$ is a non-trivial subgroup of $S_n$.
\begin{thm}[Kozlov, \cite{clmap}] % p. 198
For any $n\geq 3$, the topological space $\Delta(\overline{\Pi}_n)/S_n$ is contractible.
\end{thm}
We consider the Young subgroup $S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}$.
\begin{thm}[Donau, \cite{donau}]
\label{rdthm}
Let $n\geq3$ and $G\subset S_1\times S_{n-1}$ be a subgroup, then the topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{thm}
This leads to a general question of determining the homotopy type of $\Delta(\overline{\Pi}_n)/G$ for an arbitrary subgroup $G\subset S_n$. One might conjecture that $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres for any $n\geq 3$ and any subgroup $G\subset S_n$. But unfortunately this statement is not true as the following example will show: Let $p\geq 5$ be a prime number and let $C_p$ denote the subgroup of $S_p$ that is generated by the cycle $(1,2,\dots,p)$. Then the fundamental group of $\Delta(\overline{\Pi}_p)/C_p$ is isomorphic to $\Z/p\Z$. In particular $\Delta(\overline{\Pi}_p)/C_p$ cannot be homotopy equivalent to a wedge of spheres. A proof, which uses facts about covering spaces\footnote{See \cite[Chapter 1.3]{hatcher}}, can be found in \cite{torsion}.

@@

Let $r\in R$ and assume $G_r$ acts transitively on $C_r$. Then $G$ acts transitively on
\[
\bigcup_{g\in G}gC_r
\]
\end{prop}
\begin{proof}
We define acyclic matchings on the fibers of $\varphi$ as follows. For each $q\in Q$ we choose $r\in R$ and $g\in G$ with $gr=q$. The map $g:\varphi^{-1}(r)\longrightarrow\varphi^{-1}(q)$, which is an isomorphism of posets, induces an acyclic matching on $\varphi^{-1}(q)$. If we choose another $h\in G$ with $hr=q$, then we obtain the same matching. By Theorem \ref{patchwork} the union of these acyclic matchings is an acyclic matching which is $G$-equivariant by construction. The second statement is easy to see.
\end{proof}
\begin{figure}[ht]
\centering
\includegraphics{eqex.pdf}
\caption{A simple example: An $\Z_2$-equivariant acyclic matching composed of acyclic matchings on the fibers of $0$ and $1$. The matching pair in the fiber of $1$ is copied to the fiber of $2$. $\Z_2$ acts on both posets by reflection across the vertical line.}
\label{eqex}
\end{figure}
\begin{rem}
\label{quotmatching}
Let $G$ be a group acting on a finite regular trisp $\Delta$. Assume we have an $G$-equivariant acyclic matching $M$ on ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $C$ be the set of critical simplices. Clearly we have an action of $G$ on $C$. Let $H\subset G$ be a subgroup. Then $M/H$ is an acyclic matching on ${\cal F}(\Delta/H)\setminus\{\hat{0}\}$, where $C/H$ is the set of critical simplices. In particular, if $\Delta$ is $G$-collapsible, then $\Delta/H$ is collapsible. Furthermore if $H$ is a normal subgroup, then the acyclic matching $M/H$ is $(G/H)$-equivariant.
\end{rem}
We also have an equivariant version of the Main Theorem of Discrete Morse Theory.
\begin{thm}[Freij, \cite{freij}]
\label{freij}
Let $G$ be a finite group. Let $\Delta$ be a finite regular $G$-trisp and let $M$ be a $G$-equivariant acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then $\Delta$ is $G$-homotopy equivalent to a $G$-CW complex where the cells correspond to the critical simplices of $M$ and the action of $G$ is the same as the action on $\Delta$ restricted to the critical simplices of $M$.
\end{thm}
There exists a characterization of acyclic matchings by means of order-preserving maps with small fibers, see Definition 11.3 and Theorem 11.4 in \cite[Chapter 11]{buch}. In a similar way we can also characterize $G$-equivariant acyclic matchings by means of order-preserving $G$-maps with small fibers.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}

>>>>

\documentclass{elsarticle}
\usepackage{amsmath,amsthm}
\usepackage{bbm}
\newtheorem{df}{Definition}[section]
\newtheorem{prop}[df]{Proposition}
\newtheorem{lem}[df]{Lemma}
\newtheorem{thm}[df]{Theorem}
\newtheorem{cor}[df]{Corollary}
\newtheorem{ex}[df]{Example}
\newtheorem{rem}[df]{Remark}
\def\N{\mathbbm N}
\def\Z{\mathbbm Z}
\def\C{\mathbbm C}
\def\id{\operatorname{id}}
%\def\card{\operatorname{card}}
\journal{arXiv.org}
\begin{document}
\begin{frontmatter}
\title{A nice acyclic matching on the nerve of the partition lattice}
\author[ruelle]{Ralf Donau}
\address[ruelle]{Fachbereich Mathematik, Universit\"at Bremen, Bibliothekstra\ss e 1, 28359 Bremen, Germany}
\ead{ruelle@math.uni-bremen.de}
\begin{abstract}
The author has already proven that the space $\Delta(\Pi_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$ for all natural numbers $n\geq 3$ and all subgroups $G\subset S_1\times S_{n-1}$. We wish to construct an acyclic matching on $\Delta(\Pi_n)/G$ that allows us to give a basis of its cohomology. This is also a more elementary approach to determining the number of spheres. Furthermore we give a description of the group action by an action on the spheres. We also obtain another result that we call Equivariant Patchwork Theorem.
\end{abstract}
\begin{keyword}
Discrete Morse Theory\sep Regular trisp\sep Acyclic matching\sep Equivariant homotopy
\end{keyword}
\end{frontmatter}
\section{Introduction}
Let $n\geq 3$ and let $\Pi_n$ denote the poset consisting of all partitions of $[n]:=\{1,\dots,n\}$ ordered by refinement, such that the finer partition is the smaller partition. Let $\overline{\Pi}_n$ denote the poset obtained from $\Pi_n$ by removing both the smallest and greatest element, which are $\{\{1\},\dots,\{n\}\}$ and $\{[n]\}$, respectively. We consider $\overline{\Pi}_n$ as a category, which is acyclic, and define $\Delta(\overline{\Pi}_n)$ to the nerve of the acyclic category $\overline{\Pi}_n$, which is a regular trisp, see \cite[Chapter 10]{buch}. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way.

It is well-known that $\Delta(\overline{\Pi}_n)$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The following two theorems are results concerning the topology of the quotient $\Delta(\overline{\Pi}_n)/G$, where $G$ is a non-trivial subgroup of $S_n$.
\begin{thm}[Kozlov \cite{clmap}]
For any $n\geq 3$, the topological space $\Delta(\overline{\Pi}_n)/S_n$ is contractible.
\end{thm}
We consider the Young subgroup $S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}$.
\begin{thm}[Donau \cite{donau}]
\label{rdthm}
Let $n\geq3$ and $G\subset S_1\times S_{n-1}$ be a subgroup, then the topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{thm}
This leads to a general question of determining the homotopy type of $\Delta(\overline{\Pi}_n)/G$ for an arbitrary subgroup $G\subset S_n$. One might conjecture that $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres for any $n\geq 3$ and any subgroup $G\subset S_n$. But unfortunately this statement is not true as the following example will show: Let $p\geq 5$ be a prime number and let $C_p$ denote the subgroup of $S_p$ that is generated by the cycle $(1,2,\dots,p)$. Then the fundamental group of $\Delta(\overline{\Pi}_p)/C_p$ is isomorphic to $\Z/p\Z$. In particular $\Delta(\overline{\Pi}_p)/C_p$ cannot be homotopy equivalent to a wedge of spheres. A proof, which uses facts about covering spaces\footnote{See \cite[Chapter 1.3]{hatcher}}, can be found in \cite{torsion}.

@@

Let $r\in R$ and assume $G_r$ acts transitively on $C_r$. Then $G$ acts transitively on
\[
\bigcup_{g\in G}gC_r
\]
\end{prop}
\begin{proof}
We define acyclic matchings on the fibers of $\varphi$ as follows. For each $q\in Q$ we choose $r\in R$ and $g\in G$ with $gr=q$. The map $g:\varphi^{-1}(r)\longrightarrow\varphi^{-1}(q)$, which is an isomorphism of posets, induces an acyclic matching on $\varphi^{-1}(q)$. If we choose another $h\in G$ with $hr=q$, then we obtain the same matching. By Theorem \ref{patchwork} the union of these acyclic matchings is an acyclic matching which is $G$-equivariant by construction. The second statement is easy to see.
\end{proof}
\begin{figure}[ht]
\centering
\includegraphics{eqex.pdf}
\caption{A simple example: An $\Z_2$-equivariant acyclic matching composed of acyclic matchings on the fibers of $0$ and $1$. The matching pair in the fiber of $1$ is copied to the fiber of $2$. $\Z_2$ acts on both posets by reflection across the vertical line.}
\label{eqex}
\end{figure}
\begin{rem}
\label{quotmatching}
Let $G$ be a group acting on a finite regular trisp $\Delta$. Assume we have an $G$-equivariant acyclic matching $M$ on ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $C$ be the set of critical simplices. Clearly we have an action of $G$ on $C$. Let $H\subset G$ be a subgroup. Then $M/H$ is an acyclic matching on ${\cal F}(\Delta/H)\setminus\{\hat{0}\}$, where $C/H$ is the set of critical simplices. In particular, if $\Delta$ is $G$-collapsible, then $\Delta/H$ is collapsible. Furthermore if $H$ is a normal subgroup, then the acyclic matching $M/H$ is $(G/H)$-equivariant.
\end{rem}
We also have an equivariant version of the Main Theorem of Discrete Morse Theory.
\begin{thm}[Freij \cite{freij}]
\label{freij}
Let $G$ be a finite group. Let $\Delta$ be a finite regular $G$-trisp and let $M$ be a $G$-equivariant acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then $\Delta$ is $G$-homotopy equivalent to a $G$-CW complex where the cells correspond to the critical simplices of $M$ and the action of $G$ is the same as the action on $\Delta$ restricted to the critical simplices of $M$.
\end{thm}
There exists a characterization of acyclic matchings by means of order-preserving maps with small fibers, see Definition 11.3 and Theorem 11.4 in \cite[Chapter 11]{buch}. In a similar way we can also characterize $G$-equivariant acyclic matchings by means of order-preserving $G$-maps with small fibers.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
Now we consider the top cohomology $H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ of $\Delta(\overline{\Pi}_n)$. For each simplex $\sigma\in C_n$ we obtain a $\sigma^*\in H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ which is represented by a cocycle that maps $\sigma$ to $1$ and all other simplices to $0$. We obtain a basis $\{\sigma^*\mid\sigma\in C_n\}$ of the top cohomology which is called \emph{splitting basis} in \cite{wachs}.
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}
